Designing and modeling a browser-based distributed evolutionary computation system

### 2015-04-26 GECCO/modeling-volunteer-computing-gecco.tex
Contestando a @pacastillo
Adds reviews, fixes typos and follows reviewers's suggestions

\title{Designing and modeling a browser-based distributed evolutionary computation systems}

@@

Besides, since Amazon started selling EC2 several years ago, reliable
and scalable computing resources are also available for a low price
and on demand. Recently, Google has also refurbished its offering
lowering their prices. This means that the conjunction of free or
low-cost cloud computing engines, volunteer computing systems and
untapped capability of desktop systems can be used for creating
massive, or at least potentially massive, distributed computing
experiments. These experiments can be easily created using open-source
repository sites like GitHub\footnote{\url{http://www.github.com}} and deployed automatically to Platform as
a Service (PaaS) sites such as
Heroku\footnote{\url{https://www.heroku.com/}} or
OpenShift\footnote{\url{https://www.openshift.com/}} or even
Infraestructure as a Service (IaaS) sites such as Microsoft's Azure.

In general, it should be a requisite for volunteer computing
experiments  to be performed 
``in the open''. The fact that somebody is giving to you, basically for
free, computing resources, means that the scientist using them has to
give back; besides, people do not usually visit a resource unless they
can trust it; trust is an essential factor in the sharing of resources
such as this one, computing power \cite{van2009would}. The point of
departure for establishing this trust is releasing the source code used: all
volunteer computing platforms, from SETI@home on, have done it. The
opposite is probably the reason why many companies like PopularPower \cite{buyya2001compute}
have folded or others like CrowdProcess have decided to turn their
product to in-house computation \cite{ycombinator}: there must be a mutual relationship
of trust among the scientist and the person that is running his/her
code in the browser. As it has been mentioned in the early stages of what
was then called {\em desktop grid computing} \cite{gc:bausch}, in
fact CPU cycle selling might not make economic sense since there are
not so much demand for it and supply is quite high. However, while
potential supply is in fact huge, {\em actual} supply will depend on
the person holding it willing to actually allocate it to a particular
company selling it or a particular experiment needing it, not to
mention the fact that the experiment {\em actually} has to draw the
attention of the supplier, and thus it is better analyzed using the
Uses and Gratifications Theory \cite{severin2010communication}, that
is, taking into account {\em why} an user would do it, than using a
purely computational approach. That is why considering trust is
essential, and using free software might not be enough: Openness
has to progress from open source code to open science: releasing all
data obtained in the experiment in a repository such as GitHub,
mentioned above, and even allowing real-time access to experimental data
to users.

@@

These are best practices that have been followed in the experiments in
this paper, that first presents a new version of a platform for
distributed evolutionary algorithms (EAs) using the browser and a free (as
in free beer) backend, and second, shows the result from a
statistical point of view in order to put the basis for a model of the
metacomputer obtained by joining all volunteers and the free backend
used for the experiment. This has been one of the criterion when
choosing the different cloud services (hosting and logging, mainly):
the fact that they are free, and sustainably so
\cite{Merelo2014,jj:idc:lowcost}.

@@

% Rewrite this - JJ
In order to test the volunteer computing environment, a presentation
describing a low cost volunteer computing environment was
created. This presentation was actually delivered in several
conferences (such as IDC 2014, PPSN 2014 and FOSDEM 2015). These
results have been presented in \cite{arxiv:modeling}. In this paper,
although it uses exactly the same evolutionary algorithm client code,
the {\em presentation} has changed: it uses a single page with several
panels showing the progress of the algorithm and how many users have
been contributing to the experiment. To get users, the URL
(\url{http://nodio-jmerelo.rhcloud.com} and
\url{http://nodeo-i.herokuapp.com}) where presented, with a short
message, in Twitter and other social networks, introducing it as a
``demo for evolutionary algorithms'', sometimes in Spanish, some in
English.

@@

As stated above, the evolutionary algorithms run mainly in the
browser. The problem run is a multimodal problem called {\em l-trap},
which has been used extensively as a benchmark for evolutionary
algorithms \cite{fernandes2009using,nijssen2003analysis}. This
function counts the number of bits in a sequence of $l$ and assigns
the local maximum if it has got 0 bits and the global maximum to $l$
bits. The fitness falls into a {\em trap} as you increase the number
of bits, that is why it creates a rugged fitness landscape that is
difficult to surmount for evolutionary algorithms. Its difficulty is
increased as the number of  concatenated traps grows, turning it into
a difficult problem for evolutioanry algorithms. In
our case we have used $l=40$ traps.

This fitness is implemented as part of the NodEO evolutionary
algorithm library written in node.js \cite{DBLP:conf/gecco/GuervosVGES14}, a server-based implementation of
JavaScript, converted to a client-based version via
``browserification''. The algorithm runs a canonican evolutionary
algorithm on a population of 128, which is in fact too small for a
single node to find the solution; at least a population of 512 has
been found to be necessary to find the solution with a certain
probability. Initialization is random initialization and a canonical evolutionary algorithm
with an elite of 2 individuals is run, with crossover and then mutation applied
to all the population pool generated using 2-individual
tournament selection. The particular evolutionary
algorithm used is not so important, since we were rather focused on
the user behavior, although it is quite clear that the selective
pressure of the algorithm will have an impact in its performance and
will play for or against its asynchrony in many different and complex
ways \cite{jj:2008:PPSN}.

@@

The server program includes also additional logic for logging to
different files or services, routes to obtain information from the
server, such as the anonymized IPs that have participated in the
experiment, the ordinal number of the experiment that is being performed or the content of the chromosome pool, and code to serve the static pages
and the JavaScript/CSS
code also. The main changes to the code from the  the previous paper
\cite{arxiv:modeling} is, besides the inclusion of logging for a
better experiment analyis, the function that checks for the algorithm
termination condition and restarts the chromosome pool when
it does; it includes a configurable termination
condition that checks for the solution and empties the chromosome pool
and logs the end of the experiment. This allows us to run continuously
experiments without restarting the server by hand as we did in the
previous version. %FERGU: es esta la única innovación al paper de
                  %Arxiv? A lo mejor no deberíamos mencionarlo ni
                  %usarlo como base
% La principal innovación es que es más fácil recoger datos del
% servidor y que el despliegue es automático - JJ

@@

The experiments done with the first versions of this framework have
already been published elsewhere \cite{arxiv:modeling}. We proved that
up to 20 nodes could participate in a single run if it was
sufficiently announced; every experiment was individually announced so
more or less a certain amount of nodes were guaranteed. However, it
did not run correctly in mobile phones, design of the presentation was
not adaptive and thus missed a good amount of possible clients. In
this case the design, which has used the Kube CSS framework, was fully
adaptive and indeed proved to work with mobile phones browsers for
iPhones and Android phones. 
%
\begin{figure*}[h!tb]
  \includegraphics[width=0.90\linewidth]{img/openshift-intervals.png}
  \caption{Histogram of the intervals between two requests by the same
  IP for all experiments in the OpenShift server.}  \label{fig:intervals}
\end{figure*}
%
\begin{figure*}[h!tb]
  \includegraphics[width=0.90\linewidth]{img/openshift-intervals-IP.png}
  \caption{Boxplot of the intervals between two requests by the same
  IP, by IP, ordered by descending median.}  \label{fig:intervalIP}
\end{figure*}

At the time of this writing, 23 runs were done in OpenShift and one in  
% para trabajos futuros, estas cifras de cuántas ejecuciones se han completado 
% quizás habría que presentarlas de forma diferente. Quizás indicando el tiempo
% total de "puesta online" del sistema, cómo se publicitó para atraer gente,
% y entonces ahí contar las estadísticas de cuánta gente partició, cuántas
% ejecuciones se terminaron, etc.  [Pedro]
Heroku. The Openshift URL was mentioned more frequently than the other
so that accounts for the popularity, not any particular problem with
the algorithm in one or the other. That is why the data that we have
obtained and used below, and which is, as usual, available with a free
license, is mainly processed from the log files stored in Loggly and
in OpenShift. These 23 experiments took each one a number of hours, so
that in many cases the nodes were not accessing at the same time, but
sequentially, to the server. This means that the island model is, in
fact, behaving partly as a pool-based evolutionary algorithm
\cite{sofea:evopar2012} with some results, the best from each
generation, available to future runs. This is clearly noticed when
visualizing the algorithm: its behavior changes when it gets its first
chromosome from the pool, in many cases improving a lot.

@@

Another interesting result is that we have been able to obtain many
more different users. At the time of this writing we obtained 75
different IPs contributing CPU time to the experiment, giving an
% ¿se podría hacer en el futuro algún tipo de estudio acerca de la procedencia?
% No estoy seguro de que esa relevante, pero bueno...   [Pedro]
average of around 4 per experiment. However, there is a big variation
in the number of CPUs. There is also a much bigger range in the
performance of these clients. In order to measure that we have
computed the time between two requests from the same IP and plotted it
as an histogram in Figure \ref{fig:intervals}. In that histogram we
can see that there is an almost uniform distribution of intervals,
from the smallest taking a few milliseconds to the biggest with almost
1 second; the setup of the experiment allows it to be accessible from
a very wide range of different IPs. Even within a single IP the
variation can be relatively high, which is shown in a boxplot in
Figure \ref{fig:intervalIP}, sorted by median time. The time between
requests to the server, that is, the time to process 100 generations
of 128 individuals of an evolutionary algorithm to optimize the Trap
function has also very wide range, but even within the same IP,
inter-quartile differences of several hundred milliseconds are not
uncommon. Median for all IPs is 485 milliseconds and average is quite
close, at 496.8 ms.

@@

From the last experiment published we concluded that the nature of the
performance, which is due to the number of persons deciding to
participate in the experiment, is not totally random. There are at
least lower bounds we can count on: several computers (as few as six)
can almost always be relied on, and in some cases up to 30 can
participate in a single experiment. However, it is not clear how all
these computers contribute to the overall performance in terms of
time, although all experiments were performed until a solution was
found. In the last experiment, performed mainly in front of a live
audience using lapto computers, most of the participants in the experiment were in the same
performance tier, although a small percentage of them (which will be
around 25\%) are quite slow. In this experiment the conclusion is
different: uniform distribution of performance from quite slow to
blazing fast, with approximately 50\% slower than the average and,
conversely, 50\% faster. Our algorithm should, then, cater to a very
wide performance range so some methodology for self-adaptation should
be used on the evolutionary algorithm running in the browser. A more
precise model of the user is needed also in order to properly design
the algorithm, carrying it further away from the canonical
evolutionary algorithm we are using right now, which can only take
performance so far.

@@

This work has been supported in part by TIN2011-28627-C04-02 and
TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
SPIP2014-01437 (Direcci{\'o}n General de Tr{\'a}fico) and PYR-2014-17
GENIL project (CEI-BIOTIC Granada).

@@

\end{document}

>>>>

\title{Designing and modeling a browser-based distributed evolutionary computation system}

@@

Besides, since Amazon started selling EC2 several years ago, reliable
and scalable computing resources are also available for a low price
and on demand. Recently, Google has also refurbished its offering
lowering their prices. This means that the conjunction of free or
low-cost cloud computing engines, volunteer computing systems and
untapped capability of desktop systems can be used for creating
massive, or at least potentially massive, distributed computing
experiments. These experiments can be easily created using open-source
repository sites like GitHub\footnote{\url{http://www.github.com}} and deployed automatically to Platform as
a Service (PaaS) sites such as
Heroku\footnote{\url{https://www.heroku.com/}} or
OpenShift\footnote{\url{https://www.openshift.com/}} or even
Infrastructure as a Service (IaaS) sites such as Microsoft's Azure.

In general, it should be a requisite for volunteer computing
experiments  to be performed 
``in the open''. The fact that somebody is giving to you, basically for
free, computing resources, means that the scientist using them has to
give back; besides, people do not usually visit a resource unless they
can trust it; trust is an essential factor in the sharing of resources
such as this one, computing power \cite{van2009would}. The point of
departure for establishing this trust is releasing the source code used: all
volunteer computing platforms, from SETI@home on, have done it. The
opposite is probably the reason why many companies like PopularPower \cite{buyya2001compute}
have folded or others like CrowdProcess have decided to turn their
product to in-house computation \cite{ycombinator}: there must be a mutual relationship
of trust among the scientist and the person that is running his/her
code in the browser. As it has been mentioned in the early stages of what
was then called {\em desktop grid computing} \cite{gc:bausch}, in
fact CPU cycle selling might not make economic sense since there is
not so much demand for it and supply is quite high. However, while
potential supply is in fact huge, {\em actual} supply will depend on
the person holding it willing to actually allocate it to a particular
company selling it or a particular experiment needing it, not to
mention the fact that the experiment {\em actually} has to draw the
attention of the supplier, and thus it is better analyzed using the
Uses and Gratifications Theory \cite{severin2010communication}, that
is, taking into account {\em why} an user would do it, than using a
purely computational approach. That is why considering trust is
essential, and using free software might not be enough: Openness
has to progress from open source code to open science: releasing all
data obtained in the experiment in a repository such as GitHub,
mentioned above, and even allowing real-time access to experimental data
to users.

@@

These are best practices that have been followed in the experiments in
this paper, that first presents a new version of a platform for
distributed evolutionary algorithms (EAs) using the browser and a free (as
in free beer) backend, and second, shows the result from a
statistical point of view in order to put the basis for a model of the
meta-computer obtained by joining all volunteers and the free backend
used for the experiment. This has been one of the criterion when
choosing the different cloud services (hosting and logging, mainly):
the fact that they are free, and sustainably so
\cite{Merelo2014,jj:idc:lowcost}.

@@

% Rewrite this - JJ
In order to test the volunteer computing environment, a presentation
describing a low cost volunteer computing environment was
created. This presentation was actually delivered in several
conferences (such as IDC 2014, PPSN 2014 and FOSDEM 2015). These
results have been presented in \cite{arxiv:modeling}. In this paper,
although it uses exactly the same evolutionary algorithm client code,
the {\em presentation} has changed: it uses a single page with several
panels showing the progress of the algorithm and how many users have
been contributing to the experiment. To get users, the URL
(\url{http://nodio-jmerelo.rhcloud.com} and
\url{http://nodeo-i.herokuapp.com}) were presented, with a short
message, in Twitter and other social networks, introducing it as a
``demo for evolutionary algorithms'', sometimes in Spanish, some in
English.

@@

As stated above, the evolutionary algorithms run mainly in the
browser. The problem run is a multimodal problem called {\em l-trap},
which has been used extensively as a benchmark for evolutionary
algorithms \cite{fernandes2009using,nijssen2003analysis}. This
function counts the number of bits in a sequence of $l$ and assigns
the local maximum if it has got 0 bits and the global maximum to $l$
bits. The fitness falls into a {\em trap} as you increase the number
of bits, that is why it creates a rugged fitness landscape that is
difficult to surmount for evolutionary algorithms. Its difficulty is
increased as the number of  concatenated traps grows, turning it into
a difficult problem for evolutionary algorithms. In
our case we have used $l=40$ traps.

This fitness is implemented as part of the NodEO evolutionary
algorithm library written in node.js \cite{DBLP:conf/gecco/GuervosVGES14}, a server-based implementation of
JavaScript, converted to a client-based version via
``browserification''. The algorithm runs a Canonical evolutionary
algorithm on a population of 128, which is in fact too small for a
single node to find the solution; at least a population of 512 has
been found to be necessary to find the solution with a certain
probability. Initialization is random initialization and a canonical evolutionary algorithm
with an elite of 2 individuals is run, with crossover and then mutation applied
to all the population pool generated using 2-individual
tournament selection. The particular evolutionary
algorithm used is not so important, since we were rather focused on
the user behavior, although it is quite clear that the selective
pressure of the algorithm will have an impact in its performance and
will play for or against its asynchrony in many different and complex
ways \cite{jj:2008:PPSN}.

@@

The server program includes also additional logic for logging to
different files or services, routes to obtain information from the
server, such as the anonymized IPs that have participated in the
experiment, the ordinal number of the experiment that is being performed or the content of the chromosome pool, and code to serve the static pages
and the JavaScript/CSS
code also. The main changes to the code from the  the previous paper
\cite{arxiv:modeling} is, besides the inclusion of logging for a
better experiment analysis, the function that checks for the algorithm
termination condition and restarts the chromosome pool when
it does; it includes a configurable termination
condition that checks for the solution and empties the chromosome pool
and logs the end of the experiment. This allows us to run continuously
experiments without restarting the server by hand as we did in the
previous version. %FERGU: es esta la única innovación al paper de
                  %Arxiv? A lo mejor no deberíamos mencionarlo ni
                  %usarlo como base
% La principal innovación es que es más fácil recoger datos del
% servidor y que el despliegue es automático - JJ

@@

The experiments done with the first versions of this framework have
already been published elsewhere \cite{arxiv:modeling}. We proved that
up to 20 nodes could participate in a single run if it was
sufficiently announced; every experiment was individually announced so
more or less a certain amount of nodes were guaranteed. However, it
did not run correctly in mobile phones, design of the presentation was
not adaptive and thus missed a good amount of possible clients. In
this case the design, which has used the Kube CSS framework, was fully
adaptive and indeed proved to work with mobile phones browsers for
iPhones and Android phones. 
%
\begin{figure}[h!tb]
  \includegraphics[width=0.95\linewidth]{img/openshift-intervals.png}
  \caption{Histogram of the intervals between two requests by the same
  IP for all experiments in the OpenShift server.}  \label{fig:intervals}
\end{figure}
%
\begin{figure*}[h!tb]
  \includegraphics[width=0.90\linewidth]{img/openshift-intervals-IP.png}
  \caption{Boxplot of the intervals between two requests by the same
  IP, by IP, ordered by descending median.}  \label{fig:intervalIP}
\end{figure*}

At the time of this writing, 23 runs were done in OpenShift and one in  
% para trabajos futuros, estas cifras de cuántas ejecuciones se han completado 
% quizás habría que presentarlas de forma diferente. Quizás indicando el tiempo
% total de "puesta online" del sistema, cómo se publicitó para atraer gente,
% y entonces ahí contar las estadísticas de cuánta gente partició, cuántas
% ejecuciones se terminaron, etc.  [Pedro]
% se terminaron todas. Lo de cuanta gente por ejecución (cuantas IPs,
% en realidad) y cuanto tardaron no lo he llegado a analizar, aunque
% tengo los datos. Se deja para trabajo futuro, puedes añadirlo a un
% issue - JJ
Heroku. The Openshift URL was mentioned more frequently than the other
so that accounts for the popularity, not any particular problem with
the algorithm in one or the other. That is why the data that we have
obtained and used below, and which is, as usual, available with a free
license, is mainly processed from the log files stored in Loggly and
in OpenShift. These 23 experiments took each one a number of hours, so
that in many cases the nodes were not accessing at the same time, but
sequentially, to the server. This means that the island model is, in
fact, behaving partly as a pool-based evolutionary algorithm
\cite{sofea:evopar2012} with some results, the best from each
generation, available to future runs. This is clearly noticed when
visualizing the algorithm: its behavior changes when it gets its first
chromosome from the pool, in many cases improving a lot.

@@

Another interesting result is that we have been able to obtain many
more different users. At the time of this writing we obtained 75
different IPs contributing CPU time to the experiment, giving an
% ¿se podría hacer en el futuro algún tipo de estudio acerca de la procedencia?
% No estoy seguro de que esa relevante, pero bueno...   [Pedro]
% ¿Por país? Por referrer igual sí... También añádelo como issue - JJ
average of around 4 per experiment. However, there is a big variation
in the number of CPUs. There is also a much bigger range in the
performance of these clients. In order to measure that we have
computed the time between two requests from the same IP and plotted it
as an histogram in Figure \ref{fig:intervals}. In that histogram we
can see that there is an almost uniform distribution of intervals,
from the smallest taking a few milliseconds to the biggest with almost
1 second; the setup of the experiment allows it to be accessible from
a very wide range of different IPs. Even within a single IP the
variation can be relatively high, which is shown in a boxplot in
Figure \ref{fig:intervalIP}, sorted by median time. The time between
requests to the server, that is, the time to process 100 generations
of 128 individuals of an evolutionary algorithm to optimize the Trap
function has also very wide range, but even within the same IP,
inter-quartile differences of several hundred milliseconds are not
uncommon. Median for all IPs is 485 milliseconds and average is quite
close, at 496.8 ms.

@@

From the last experiment published we concluded that the nature of the
performance, which is due to the number of persons deciding to
participate in the experiment, is not totally random. There are at
least lower bounds we can count on: several computers (as few as six)
can almost always be relied on, and in some cases up to 30 can
participate in a single experiment. However, it is not clear how all
these computers contribute to the overall performance in terms of
time, although all experiments were performed until a solution was
found. In the last experiment, performed mainly in front of a live
audience using laptop computers, most of the participants in the experiment were in the same
performance tier, although a small percentage of them (which will be
around 25\%) are quite slow. In this experiment the conclusion is
different: uniform distribution of performance from quite slow to
blazing fast, with approximately 50\% slower than the average and,
conversely, 50\% faster. Our algorithm should, then, cater to a very
wide performance range so some methodology for self-adaptation should
be used on the evolutionary algorithm running in the browser. A more
precise model of the user is needed also in order to properly design
the algorithm, carrying it further away from the canonical
evolutionary algorithm we are using right now, which can only take
performance so far.

@@

This work has been supported in part by TIN2011-28627-C04-02 and
TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
SPIP2014-01437 (Direcci{\'o}n General de Tr{\'a}fico) and PYR-2014-17
GENIL project (CEI-BIOTIC Granada). We would also like to thank the
anonymous reviewers of this paper who have really helped us to improve
this paper (and our work) with their suggestions.

@@

\end{document}
%%% Local Variables:
%%% ispell-local-dictionary: "english"
%%% End:
